---
title: "Towards a mechanistic understanding of the drivers of plant-pollinator interactions"
format:
  pdf:
    include-in-header: 
      text: |
        \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
        \usepackage{caption}
        \captionsetup[figure]{labelformat=empty}
        \captionsetup[table]{labelformat=empty}
        \usepackage{float}
        \floatplacement{figure}{H}
        
bibliography: References.bib
csl: apa.csl
latex-tinytex: false
link-citations: TRUE
linkcolor: RoyalBlue
urlcolor: RoyalBlue
---
```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,}
#Load libraries
#Load libraries
library(dplyr) #data organization
library(vegan) #nmds
library(tidyr) #pivot wider
library(ggplot2) #plotting
library(viridis) #plotting
library(patchwork) #plotting
library(tibble) #cols to rows

```


\vspace{-2cm}

Jose B. Lanuza^1,2^, Alfonso Allen-Perkins^3^,  Henriette F. Morgenroth^1^,  Raymond Umazekabiri^1^, Will Glenny^1^, Marc Hoffmann^1^, Panagiotis Theodorou^1,4^, Robert Paxton^1,4^, Isabell Hensen^1,4^,  Robert Rauschkolb^1,5^, Christine Römermann^1,5^, Oliver Schweiger ^1,7^, Martin Freiberg^1,6^, Tiffany Knight^1^

*Affiliations*

1. German Centre for Integrative Biodiversity Research (iDiv) Halle-JenaLeipzig, Puschstr. 4, 04103 Leipzig, Germany 

2. Mediterranean Institute for Advanced Studies (IMEDEA, UIB-CSIC), Mallorca, Balearic Islands, Spain

3. Departamento de Ingeniería Eléctrica, Electrónica, Automática y Física Aplicada, ETSIDI, Universidad Politécnica de Madrid, 28040 Madrid, Spain.


4. Institute for Biology, Martin-Luther-UniversityHalle-Wittenberg, Hoher Weg 8, 06120 Halle (Saale), Germany 

5. Institute of Ecology and Evolution with Herbarium Haussknecht and Botanical Garden, Friedrich Schiller University Jena, Jena, Germany 

6. Systematic Botany and Functional Biodiversity/Botanical Garden, Institute of Biology, Leipzig University, Johannisallee 21-23, 04103, Leipzig, Germany 

7. UFZ-Helmholtz Centre for Environmental Research, Department of Community Ecology, 06120 Halle, Germany

# Abstract

Flowering plants exhibit an astonishing diversity of reproductive structures, which  shapes how they interact with their pollinators. However, to what extent reproductive traits or other drivers like phenology and species abundances shape plant-pollinator interactions is still an open question. To disentangle the relative effects of species phenology, abundances and traits in plant-pollinator interactions, we compared pollinator visitation rates, diet breath, specialization from plant species that vary in reproductive traits, phenology and abundances. To do this, we observed plant-pollinator interactions for a whole flowering season in three different botanical gardens located in an urban setting. We anticipate that the ecological role of the different plant and pollinator species are largely driven by their phenological overlap and abundance, and only certain taxonomic groups are strongly influenced by trait matching processes. Our results highlight the relevance of considering multiple mechanisms to understand plant-pollinator interactions.


# Introduction

1st paragrap 

Flowering plants, pollinators, and their interactions form complex networks that are fundamental to the architecture of biodiversity. Each of these interactions can be shaped by distinct ecological drivers, including species' local abundances, traits, and activity periods. However, our ability to predict and understand ecological patterns from these networks is strongly influenced by the spatiotemporal dynamics of the community and sampling effort. For instance, a species may be classified as a specialist due to under-sampling, rarity, or limited availability of floral resources. Thus, there is a need to disentangle sampling artifacts due to limited data availability from the real ecology of the species. 



2nd paragraph 

Highlight the importance of abundance:

There is consensus that including abundances in our view of ecological processes shaping network patterns yields consistently better results [REFS, eg., annual reviews]. However, most ecological studies... 

Highlight the importance of phenology:
Highlight the importance of traits/phylogeny:




Last paragraph, introduce what we test!

By conducting this experiment in botanical gardens, we have as objective reducing the "noise" associated with the context-dependent nature of ecological interactions [REF].

What use and what temporal resolution of networks are best for improving our predictive understanding of the structure and dynamics of ecological communities? by Caradonna and Waser 2020


Explore how traits, phenology and abundances shape species interactions in three botanical gardens from Germany at different levels of temporal complexity 

Pending tasks:

Question 1:

- Explore relative contribution of ecological drivers (abundances, phenology, traits, phylogeny (plants and bees)).

- Associated with this, how increasing sampling impact our view of those processes.

- Temporal grain, how does it affect network patterns. Relative contribution of temporality to network stucture or patterns (e.g., day 1, only adds 1 spp versus day 10 that adds 30). 



# Methods

## Study site and sampling method

Plant-pollinator interactions were monitored during the entire flowering season of 2023 in the botanical gardens of Halle, Jena, and Leipzig (Germany). These gardens belong to PhenObs initiative [@nordt2021], which consists of a network of botanical gardens across the Northern Hemisphere that records phenological observations on a large set of herbaceous species. The objective of PhenObs is to evaluate the impacts of climate change on species performance [e.g.\, @sporbert2022]. In this study, plant-pollinator interactions were recorded for a subset of species from the PhenObs project within these three gardens. Each garden was visited weekly, and plant-pollinator interactions were sampled during 15 mins on each focal plant.  On each sampling day, the different focal plants were sampled in a random order and only pollinator individuals that contacted the reproductive organs of the plant were recorded as interacting species. In addition, to create a comprehensive inventory of pollinators and their interactions within the gardens, 10 random sites were selected each sampling day, and plant-pollinator interactions were monitored for three minutes at each random site, resulting in a total of 30 minutes per sampling day and garden. Therefore, two sampling methodologies were implemented, one that recorded interactions on PhenObs plants, and another that consisted of random stops within the garden, with the specific goal of describing all potential pollinators present in the garden.

```{r ,echo=FALSE, message=FALSE, cache=FALSE, warning="FALSE"}

# Load libraries
library(dplyr)

# Load data
raw_data = readRDS("../Data/Working_files/interaction_data.rds")
phenobs_spp = readRDS("../Data/Working_files/phenobs_spp.rds")

# Number of phenobs species
phenobs_total = length(phenobs_spp)

# Prepare data
int_data = raw_data %>%
  filter(Plant_rank == "SPECIES") %>%
  select(Plant, 
         Plant_accepted_name, 
         Pollinator_accepted_name, 
         Botanical_garden, 
         Sampling,
         Plant_order,
         Plant_family,
         Plant_genus) %>%
  rename(Plants = Plant_accepted_name,
         Pollinators = Pollinator_accepted_name) %>%
  distinct()

# Check unique number of PhenObs species by garden
phenobs_by_garden = int_data %>% 
filter(Plant %in% phenobs_spp) %>% 
select(Botanical_garden, Plant) %>% 
group_by(Botanical_garden) %>% 
summarise(Phenobs_spp = n_distinct(Plant))

# PhenObs species by garden
phenobs_h = phenobs_by_garden$Phenobs_spp[1]
phenobs_j = phenobs_by_garden$Phenobs_spp[2]
phenobs_l = phenobs_by_garden$Phenobs_spp[3]

# Check unique number of Non-PhenObs species by garden
phenobs_by_garden1 = int_data %>% 
filter(Sampling == "Random_census") %>% 
select(Botanical_garden, Plant) %>% 
group_by(Botanical_garden) %>% 
summarise(Non_phenobs_spp = n_distinct(Plant))

# PhenObs species by garden
non_phenobs_h = phenobs_by_garden1$Non_phenobs_spp[1]
non_phenobs_j = phenobs_by_garden1$Non_phenobs_spp[2]
non_phenobs_l = phenobs_by_garden1$Non_phenobs_spp[3]

# Check total number of plant orders within PhenObs
phenobs_orders = int_data %>% 
  filter(Plant %in% phenobs_spp) %>% 
  distinct(Plant_order) %>% 
  nrow()

# Check total number of plant families within PhenObs
phenobs_families = int_data %>% 
  filter(Plant %in% phenobs_spp) %>% 
  distinct(Plant_family) %>% 
  nrow()

# Check total number of plant genera within PhenObs
phenobs_genera = int_data %>% 
  filter(Plant %in% phenobs_spp) %>% 
  distinct(Plant_genus) %>% 
  nrow()

# Calculate total time per garden for PhenObs plants
time_per_garden = raw_data %>% 
  filter(Sampling == "Focal") %>% 
  select(Botanical_garden, Date, Total_time_species, Plant) %>% 
  distinct() %>% 
  group_by(Botanical_garden) %>% 
  summarise(Total_time = sum(Total_time_species) / 60)

halle_hours = round(time_per_garden$Total_time[1], 1)
jena_hours = round(time_per_garden$Total_time[2], 1)
leipzig_hours = round(time_per_garden$Total_time[3],1)


# Calculate total time per garden for NON-PhenObs plants
time_per_garden1 = raw_data %>% 
  filter(Sampling == "Random_census") %>% 
  select(Botanical_garden, Date, Total_time_species, Plant) %>% 
  distinct() %>% 
  group_by(Botanical_garden) %>% 
  summarise(Total_time = sum(Total_time_species) / 60)

halle_hours_random = round(time_per_garden1$Total_time[1], 1)
jena_hours_random = round(time_per_garden1$Total_time[2], 1)
leipzig_hours_random = round(time_per_garden1$Total_time[3],1)


# Percentage of species with capitulum
capitulum_percent = int_data = raw_data %>%
  filter(Plant_rank == "SPECIES") %>%
  #filter(Sampling == "Focal") %>% 
  #filter(Plant %in% phenobs_spp) %>% 
  select(Plant, 
         Capitulum) %>% 
  distinct() %>% 
  group_by(Capitulum) %>% 
  summarise(n_plants = length(Plant) / nrow(.) * 100)

non_capitulum_percent = capitulum_percent %>% 
  filter(Capitulum == "No") %>% 
  pull(n_plants) %>% 
  round(1)

capitulum_percent = capitulum_percent %>% 
  filter(Capitulum == "Yes") %>% 
  pull(n_plants) %>% 
  round(1)


```

## Sampling effort and plant taxonomic coverage

This study monitored a total of `r phenobs_total` focal plant species that belonged to PhenObs initiative, which accounted for `r phenobs_genera` different genera, `r phenobs_families` families and `r phenobs_orders` different orders. In total, we sampled pollinators on `r phenobs_h` plants species in Halle (`r halle_hours` hours), `r phenobs_j` in Jena (`r jena_hours` hours) and `r phenobs_l` in Leipzig (`r leipzig_hours` hours), and only 12 species were shared across the three gardens. In addition, with the random stops across the garden we sampled a total of XXXX species that comprised XXXX genera, XXXXX families and XXXXX orders. This study accounted for a total of `r non_phenobs_h` plant species in Halle (`r halle_hours_random` hours), `r non_phenobs_j` in Jena (`r jena_hours_random` hours), and `r non_phenobs_l` in Leipzig (`r leipzig_hours_random` hours). Explain accumulation curves here!!!

## Data description

To evaluate the different mechanisms that drive plant-pollinator interactions, detailed information on abundance, phenology and species traits for both plants and pollinators was recorded as follows: 

(i) Floral abundance was annotated for each focal plant species, and given that botanical gardens are a semi-control setting with relatively low abundance of pollinators, we were also able to annotate the different number of pollinator individuals on each sampled plant when documenting interactions. Despite not being fully independent of interactions, this approach offers a more realistic estimate of pollinator abundance than the standard method of estimating abundances from interactions, which can bias our understanding of the mechanisms driving network structure [@bluthgen2006; @vizentin2014]. Note that flowering species with tiny densely packed flowers were considered at inflorescence level (`r capitulum_percent`% of the total flowering species considered in this study). 


(ii) Given that plant phenology was already monitored on the PhenObs project, we have weekly phenological records with flowering percent for all focal plants. For 10 species with missing phenology data, we conducted generative additive models (GAM) with our records of flower number to build the phenology on the same sampling dates as the other species by using the *mgcv* package. Unlike plants, pollinator phenology relies on ‘stochastic’ interaction observations that may not represent their complete flying period. For this reason, we opted to complement our observational data from the Global Biodiversity Information Facility (GBIF). Germany is one of the countries with highest records of occurrence data [@rocha2021], and in particular, their cities (e.g., botanical gardens) contain most of those records [@sweet2022]. Thus, we opted to extract occurrence records around the three cities for the different pollinator species to complement the different pollinator phenologies. We first created a centroid within the 3 cities, and then, created a rectangular shape around it that expanded approximately 1000 kms in longitude and 500 kms in latitude (Figure S1). We used this shape because we expect environmental variables to be more consistent and phenology to be more similar across a longitudinal range than a latitudinal one. Thus, our download resulted in 2,933,772 occurrence records (https://doi.org/10.15468/dl.k5cnuf).

(iii) Plant reproductive traits were selected following the main traits driving the trait spectrum of flowering plants as shown in @lanuza2023 (i.e., flower number, plant height, style length, flower size, ovule number and selfing). In addition, we also included phenological traits (i.e., flower lifespan and flowering length), floral rewards (nectar volume and number of pollen grains per flower) and seed weight as a measurement of the reproductive investment. Pollinator traits included intertegular distance, body length and proboscis length. Both intertegular distance and body length were measured from images using ImageJ, and proboscis length was extrapolated for bee species by using the pollimetry package [@kendall2019]. Non-bee species were complemented with proboscis values from the literature and images when possible. 

## Plant and pollinator phylogeny


*Phylogeny... Plant phylogeny is easy, bee phylogeny too... 

*compute phenology matrices: presence absence and probability ones based by calculating the probability of observing both if they are independent events divided by total number of sampling days


CONCEPTUAL GRAPH. **Figure 1**

## Analysess

To unravel the role of time in our understanding of the cological drivers of plant-pollinator networks, this study analyses the relative importance of abundance, phenology and traits at the whole season level, by dividing in early-mid-and late season and by analysing at the sampling day level. 


# Results

PROVIDE: Percentage of forbidden links due to phenological mistmatches by garden.







# References

<div id="refs"></div>


# Supporting information

Figure S1: Some sort of combination of the code below and the graph of location from the poster

```{r ,echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, fig.height=10,fig.width=10, fig.cap = "\\textbf{Figure S1}."}

##Load data
#centroid = readRDS("../Data/Working_files/figs1_centroid.rds")
#oc_2023 = readRDS("../Data/Working_files/oc_2023.rds")
#rectangle = readRDS("../Data/Working_files/rectangle.rds")
#
#
#
#library(sf)
## Convert points to sf
#world <- map_data("world")
#ggplot() +
#  geom_map(
#    data = world,
#    map = world,
#    aes(long, lat, map_id = region),
#    color = "white",
#    fill = "gray90",
#    size = 0.1
#  ) +
#  ylim(0, 70) +
#  geom_point(
#    data = oc_2023,
#    aes(decimalLongitude, decimalLatitude),
#    color = "#0072B2",   # soft blue for points
#    alpha = 0.6,
#    size = 1.2
#  ) +
#  geom_point(
#    data = centroid,
#    aes(Longitude, Latitude),
#    color = "black",
#    size = 3
#  ) +
#  geom_sf(
#    data = rectangle,
#    colour = "red",
#    fill = NA,
#    linewidth = 1.2
#  ) +
#  coord_sf(
#    xlim = c(-15, 44),
#    ylim = c(33, 73),
#    expand = FALSE
#  ) +
#  theme_minimal() +
#  theme(
#    panel.background = element_rect(fill = "aliceblue"),  # light blue #ocean
#    panel.grid = element_line(color = "white"),  # faint grid
#    axis.text = element_text(size = 10),
#    axis.title = element_blank()
#  )
#
```


