---
title: "Towards a mechanistic understanding of the drivers of plant-pollinator interactions"
format:
  pdf:
    include-in-header: 
      text: |
        \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
        \usepackage{caption}
        \captionsetup[figure]{labelformat=empty}
        \captionsetup[table]{labelformat=empty}
        \usepackage{float}
        \floatplacement{figure}{H}
        
bibliography: References.bib
csl: apa.csl
latex-tinytex: false
link-citations: TRUE
linkcolor: RoyalBlue
urlcolor: RoyalBlue
---
```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,}
#Load libraries
#Load libraries
library(dplyr) #data organization
library(vegan) #nmds
library(tidyr) #pivot wider
library(ggplot2) #plotting
library(viridis) #plotting
library(patchwork) #plotting
library(tibble) #cols to rows

```


\vspace{-2cm}

Jose B. Lanuza^1,2^, Alfonso Allen-Perkins^3^,  Henriette F. Morgenroth^1^,  Raymond Umazekabiri^1^, Will Glenny^1^, Marc Hoffmann^1^, Panagiotis Theodorou^1,4^, Robert Paxton^1,4^, Isabell Hensen^1,4^,  Robert Rauschkolb^1,5^, Christine Römermann^1,5^, Oliver Schweiger ^1,7^, Martin Freiberg^1,6^, Tiffany Knight^1^

*Affiliations*

1. German Centre for Integrative Biodiversity Research (iDiv) Halle-JenaLeipzig, Puschstr. 4, 04103 Leipzig, Germany 

2. Mediterranean Institute for Advanced Studies (IMEDEA, UIB-CSIC), Mallorca, Balearic Islands, Spain

3. Departamento de Ingeniería Eléctrica, Electrónica, Automática y Física Aplicada, ETSIDI, Universidad Politécnica de Madrid, 28040 Madrid, Spain.


4. Institute for Biology, Martin-Luther-UniversityHalle-Wittenberg, Hoher Weg 8, 06120 Halle (Saale), Germany 

5. Institute of Ecology and Evolution with Herbarium Haussknecht and Botanical Garden, Friedrich Schiller University Jena, Jena, Germany 

6. Systematic Botany and Functional Biodiversity/Botanical Garden, Institute of Biology, Leipzig University, Johannisallee 21-23, 04103, Leipzig, Germany 

7. UFZ-Helmholtz Centre for Environmental Research, Department of Community Ecology, 06120 Halle, Germany

# Abstract

Flowering plants exhibit an astonishing diversity of reproductive structures, which  shapes how they interact with their pollinators. However, to what extent reproductive traits or other drivers like phenology and species abundances shape plant-pollinator interactions is still an open question. To disentangle the relative effects of species phenology, abundances and traits in plant-pollinator interactions, we compared pollinator visitation rates, diet breath, specialization from plant species that vary in reproductive traits, phenology and abundances. To do this, we observed plant-pollinator interactions for a whole flowering season in three different botanical gardens located in an urban setting. We anticipate that the ecological role of the different plant and pollinator species are largely driven by their phenological overlap and abundance, and only certain taxonomic groups are strongly influenced by trait matching processes. Our results highlight the relevance of considering multiple mechanisms to understand plant-pollinator interactions.


Introduction

1st paragrap 

General introductory paragraph

Ecological networks... 
Introduce drivers
End with limitations and state of the art, temporal and spatial components, sampling biases and data limitations from rarity.

2nd paragraph 

Highlight the importance of abundance:

There is consensus that including abundances in our view of ecological processes shaping network patterns yields consistently better results [REFS, eg., annual reviews]. However, most ecological studies... 

Highlight the importance of phenology:
Highlight the importance of traits/phylogeny:




Last paragraph, introduce what we test!

By conducting this experiment in botanical gardens, we have as objective reducing the "noise" associated with the context-dependent nature of ecological interactions [REF].



# Main objective

Explore how traits, phenology and abundances shape species interactions in three botanical gardens from Germany. 

Pending tasks:

Question 1:

- Explore relative contribution of ecological drivers (abundances, phenology, traits, phylogeny (plants and bees)).

      - Pollinators: conduct rarefaction on our data and compare phenobs vs random stops vs all garden  

- Associated with this, how increasing sampling impact our view of those processes.

- Temporal grain, how does it affect network patterns. Relative contribution of temporality to network stucture or patterns (e.g., day 1, only adds 1 spp versus day 10 that adds 30). 



# Methods

## Study site and sampling method

Plant-pollinator interactions were monitored during the entire flowering season of 2023 in the botanical gardens of Halle, Jena, and Leipzig (Germany). These gardens belong to PhenObs initiative [@nordt2021], which consists of a network of botanical gardens across the Northern Hemisphere that records phenological observations on a large set of herbaceous species. The objective of PhenObs is to evaluate the impacts of climate change on species performance [e.g.\, @sporbert2022]. In this study, plant-pollinator interactions were recorded for a subset of species from the PhenObs project within these three gardens. Each garden was visited weekly, and plant-pollinator interactions were sampled during 15 mins on each focal plant. This study monitored a total of 83 focal plant species (42 in Halle, 43 in Jena and 47 in Leipzig), and a total of 12 were shared across the three gardens. These different focal plants were sampled at random order each sampling day at each garden. In addition, to create a comprehensive inventory of pollinators and their interactions within the gardens, 10 random sites were selected, and plant-pollinator interactions were monitored for three minutes at each random location on each garden. Therefore, two sampling methodologies were implemented, one that recorded interactions on PhenObs plants, and another that consisted of random stops within the garden, with the specific goal of describing all potential pollinators present in the garden.

## Data description

To evaluate the different mechanisms that drive plant-pollinator interactions, detailed information on abundance, phenology and species traits for both plants and pollinators was recorded as follows: 

(i) Floral abundance was annotated for each focal plant species, and given that botanical gardens are a semi-control setting with relatively low abundance of pollinators, we were also able to annotate the different number of pollinator individuals on each sampled plant when documenting interactions. Despite not being fully independent of interactions, this approach offers a more realistic estimate of pollinator abundance than the standard method of estimating abundances from interactions, which can bias our understanding of the mechanisms driving network structure [@bluthgen2006; @vizentin2014].


Copy paste from Herrera 2020, adapt to our work: 

Only taxa whose individuals contacted anthers or stigmas, or had visible pollen grains on body surfaces, are considered as pollinators in this study.

Some species had tiny flowers densely packed into compact inflorescences (e.g., Apiaceae, Asteraceae, Dipsacaceae; N = 53 species, Table 1), which rendered impractical assessing the number of elemental florets visited by pollinators. In these cases, the number of inflorescences available per patch and visited per census was counted rather than individual flowers, and visitation probabilities (see Data analysis) thus actually refer to inflorescences. For simplicity, I will refer to visitation to both single flowers and inflorescences as “flower visitation.” In some analyses, however, the two types of visitation units will be considered as levels of the discrete variable “visitation unit,” one of the predictors of pollinator visitation considered here (“single flowers” vs. “flower packets”


(ii) Given that plant phenology was already monitored on the PhenObs project, we have weekly phenological records with flowering percent for all focal plants. For 10 species with missing phenology data, we conducted generative additive models (GAM) with our records of flower number to build the phenology on the same sampling dates as the other species by using the *mgcv* package. Unlike plants, pollinator phenology relies on ‘stochastic’ interaction observations that may not represent their complete flying period. For this reason, we opted to complement our observational data from the Global Biodiversity Information Facility (GBIF). Germany is one of the countries with highest records of occurrence data [@rocha2021], and in particular, their cities (e.g., botanical gardens) contain most of those records [@sweet2022]. Thus, we opted to extract occurrence records around the three cities for the different pollinator species to complement the different pollinator phenologies. We first created a centroid within the 3 cities, and then, created a rectangular shape around it that expanded approximately 1000 kms in longitude and 500 kms in latitude (Figure S1). We used this shape because we expect environmental variables to be more consistent and phenology to be more similar across a longitudinal range than a latitudinal one. Thus, our download resulted in 2,933,772 occurrence records (https://doi.org/10.15468/dl.k5cnuf).

(iii) Plant reproductive traits were selected following the main traits driving the trait spectrum of flowering plants as shown in @lanuza2023 (i.e., flower number, plant height, style length, flower size, ovule number and selfing). In addition, we also included phenological traits (i.e., flower lifespan and flowering length), floral rewards (nectar volume and number of pollen grains per flower) and seed weight as a measurement of the reproductive investment. Pollinator traits included intertegular distance, body length and proboscis length. Both intertegular distance and body length were measured from images using ImageJ, and proboscis length was extrapolated for bee species by using the pollimetry package [@kendall2019]. Non-bee species were complemented with proboscis values from the literature and images when possible. 

*Phylogeny... Plant phylogeny is easy, bee phylogeny too... 

*I think this article may need some sort of conceptual figure

*compute phenology matrices: presence absence and probability ones based by calculating the probability of observing both if they are independent events divided by total number of sampling days

## Analysis

Compare matrices presence/absence|visitation rate ~ abundance/phenology/traits

Check abundances (FLORAL ABUNDANCES)

# Results

Compare differences between phenobs and random stops across gardens (identity, number of species and interactions by sampling date or sampled species). Number of species of plant and pollinators on each and percentage of plant-pollinator interactions and pollinator species captured by the PhenObs species.  

*Rarefy by garden: phenobs, random stops and all. 

PROVIDE: Percentage of forbidden links due to phenological mistmatches by garden.







# References

<div id="refs"></div>


# Supporting information

Figure S1: Some sort of combination of the code below and the graph of location from the poster

```{r ,echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, fig.height=10,fig.width=10, fig.cap = "\\textbf{Figure S1}."}

##Load data
#centroid = readRDS("../Data/Working_files/figs1_centroid.rds")
#oc_2023 = readRDS("../Data/Working_files/oc_2023.rds")
#rectangle = readRDS("../Data/Working_files/rectangle.rds")
#
#
#
#library(sf)
## Convert points to sf
#world <- map_data("world")
#ggplot() +
#  geom_map(
#    data = world,
#    map = world,
#    aes(long, lat, map_id = region),
#    color = "white",
#    fill = "gray90",
#    size = 0.1
#  ) +
#  ylim(0, 70) +
#  geom_point(
#    data = oc_2023,
#    aes(decimalLongitude, decimalLatitude),
#    color = "#0072B2",   # soft blue for points
#    alpha = 0.6,
#    size = 1.2
#  ) +
#  geom_point(
#    data = centroid,
#    aes(Longitude, Latitude),
#    color = "black",
#    size = 3
#  ) +
#  geom_sf(
#    data = rectangle,
#    colour = "red",
#    fill = NA,
#    linewidth = 1.2
#  ) +
#  coord_sf(
#    xlim = c(-15, 44),
#    ylim = c(33, 73),
#    expand = FALSE
#  ) +
#  theme_minimal() +
#  theme(
#    panel.background = element_rect(fill = "aliceblue"),  # light blue #ocean
#    panel.grid = element_line(color = "white"),  # faint grid
#    axis.text = element_text(size = 10),
#    axis.title = element_blank()
#  )
#
```


