---
format:
  pdf:
    include-in-header: 
      text: |
        \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
        \usepackage{caption}
        \captionsetup[figure]{labelformat=empty}
        \captionsetup[table]{labelformat=empty}
        \usepackage{float}
        \floatplacement{figure}{H}
     
       
bibliography: References.bib
csl: apa.csl
latex-tinytex: false
link-citations: TRUE
linkcolor: RoyalBlue
urlcolor: RoyalBlue
---
```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,}
#Load libraries
#Load libraries
library(dplyr) #data organization
library(vegan) #nmds
library(tidyr) #pivot wider
library(ggplot2) #plotting
library(viridis) #plotting
library(patchwork) #plotting
library(tibble) #cols to rows

```


\begin{center}
\LARGE\textbf{Towards a mechanistic understanding of the drivers of plant-pollinator interactions}
\end{center}


Jose B. Lanuza^1,2^, Alfonso Allen-Perkins^3^,  Henriette F. Morgenroth^1^,  Raymond Umazekabiri^1^, Will Glenny^1^, Anna Traveset^2^, Marc Hoffmann^1^, Panagiotis Theodorou^1,4^, Robert Paxton^1,4^, Isabell Hensen^1,4^,  Robert Rauschkolb^1,5^, Christine Römermann^1,5^, Oliver Schweiger ^1,7^, Martin Freiberg^1,6^, Tiffany Knight^1^


\small

*Affiliations*

1. German Centre for Integrative Biodiversity Research (iDiv) Halle-JenaLeipzig, Puschstr. 4, 04103 Leipzig, Germany 

2. Mediterranean Institute for Advanced Studies (IMEDEA, UIB-CSIC), Mallorca, Balearic Islands, Spain

3. Departamento de Ingeniería Eléctrica, Electrónica, Automática y Física Aplicada, ETSIDI, Universidad Politécnica de Madrid, 28040 Madrid, Spain.


4. Institute for Biology, Martin-Luther-UniversityHalle-Wittenberg, Hoher Weg 8, 06120 Halle (Saale), Germany 

5. Institute of Ecology and Evolution with Herbarium Haussknecht and Botanical Garden, Friedrich Schiller University Jena, Jena, Germany 

6. Systematic Botany and Functional Biodiversity/Botanical Garden, Institute of Biology, Leipzig University, Johannisallee 21-23, 04103, Leipzig, Germany 

7. UFZ-Helmholtz Centre for Environmental Research, Department of Community Ecology, 06120 Halle, Germany

\normalsize

# Abstract


Flowering plants exhibit an astonishing diversity of reproductive structures, which  shapes how they interact with their pollinators. However, to what extent reproductive traits or other drivers like phenology and species abundances shape plant-pollinator interactions is still an open question. To disentangle the relative effects of species phenology, abundances and traits in plant-pollinator interactions, we compared pollinator visitation rates, diet breath, specialization from plant species that vary in reproductive traits, phenology and abundances. To do this, we observed plant-pollinator interactions for a whole flowering season in three different botanical gardens located in an urban setting. We anticipate that the ecological role of the different plant and pollinator species are largely driven by their phenological overlap and abundance, and only certain taxonomic groups are strongly influenced by trait matching processes. Our results highlight the relevance of considering multiple mechanisms to understand plant-pollinator interactions.


# Introduction

Flowering plants, pollinators, and their interactions form complex networks that are fundamental to the architecture of biodiversity. The structure and composition of these networks exhibit strong spatio-temporal variation, shaped by species attributes and environmental conditions [@burkle2011; @tylianakis2017]. While ecological networks provide insights into the ecological processes that govern community assembly [@bascompte2007; @guimaraes2020], documenting interactions remains extremely labour-intensive [@chacoff2012]. Moreover, the resulting network structures are also highly dependent on sampling effort and methodology [@jordano2016]. At the same time, interactions are often summarised into single metrics and analysed with a diverse array of statistical methods, each offering potentially distinct perspectives on the ecological system [@delmas2019; @peralta2024]. Despite pronounced temporal dynamics, networks are often treated as static entities [@poisot2015; @caradonna2017], requiring the aggregation or simplification of relevant information when evaluating ecological patterns. The complexity of plant-pollinator networks, together with the limitations of current analytical practices, can obscure rather than clarify our view of ecological systems. Addressing these challenges requires systematic approaches that integrate multiple layers of network complexity to improve both understanding and predictive power [@dormann2017; @bluthgen2024].

There is strong evidence that species phenology, abundances, and traits are key ecological drivers shaping species interactions and network structure [@vazquez2009; @kaiser2014], yet most studies lack the appropriate data to assess their combined influence [@vazquez2009; @olito2015]. Phenology is considered one of the main determinants of network structure, as temporal overlap between species constrains the pool of potential interactions [@encinas2012; @benadi2014], with approximately one-third of all potential interactions estimated to be impossible due to phenological mismatches [@olesen2011]. When two species overlap in time, the probability that they interact is generally considered a function of their relative abundances and species traits [@olito2015; @stang2007], along with other abiotic and biotic factors (e.g. temperature and species competition). While abundance influences both species’ roles [@fort2016; @simmons2019] and overall network structure [@vazquez2009; @canard2014], independent estimates are rarely available and are often inferred from interaction data [@bluthgen2024].

Traits, in turn, provide mechanistic insights, as they constrain the range of possible partners and help define species’ ecological roles within networks [@bartomeus2016; @dehling2016; @coux2016]. For instance, interaction probability often peaks when pollinator proboscis length matches floral tube depth [@klumpers2019], yet imperfect matches can still allow interactions. Although multiple traits often act simultaneously, capturing their multidimensional nature remains challenging [@eklof2013; @lanuza2023], though phylogenetic approaches may help integrate such complexity. Further, most pollinators tend to be generalist, and the relative importance of traits is expected to decline with increasing generalism [@peralta2020]. The strong context-dependency of these different ecological drivers constrains predictive power and highlights the need for high-quality data to evaluate multiple scenarios across taxonomic groups and spatiotemporal scales.

Recent evidence suggests that distinct temporal scales can provide different views of the structure and processes that shape plant-pollinator interactions [@miele2020; @schwarz2020]. Species and interactions are highly dynamic, shifting markedly between years [@petanidou2008; @chacoff2018], across seasons [@vizentin2016; @souza2018], and even within days [@schwarz2021]. At fine scales, interactions are shaped by physiological responses to abiotic factors (e.g., temperature, light, precipitation) that determine activity periods, whereas variation across days or weeks also reflects phenological succession and changes in the species pool [@caradonna2017; @bramon2020]. Although some of this variation arises from natural dynamics, global change is expected to disrupt and intensify temporal fluctuations [@kharouba2018; @duchenne2020]. While aggregated snapshots of communities are valuable for describing structural and ecological properties at broad scales [@bascompte2007], the relative impact of ecological drivers may vary with temporal scale, underscoring the need to account for temporal dynamics to fully understand plant-pollinator interactions.

Although there is no consensus on the range of tools used to analyze ecological networks, combining approaches can provide complementary insights [@delmas2019; @bluthgen2024; @dormann2025]. Structural metrics are often compared with synthetic networks or null models built under different ecological constraints to test for significance, or across empirical networks to assess variation among systems [@vazquez2004; @santamaria2007; @marjakangas2022], a strategy widely applied to evaluate properties such as nestedness [@bascompte2003; @ulrich2007; @payrato2019] and modularity [@olesen2007; @dormann2013]. Other statistical approaches can complement structural analyses by explicitly assessing the influence of biotic and abiotic drivers. Common approaches include matrix comparisons such as Mantel or Procrustes tests [@burkle2011; @dehling2016] and conventional models such as generalized linear models [@fornoff2017; @classen2020; @herrera2020], which allow more refined evaluations of the impact of biotic and abiotic drivers on species interactions. In addition, machine learning approaches provide more flexible alternatives that can capture nonlinear relationships and often outperform conventional methods. [@pichler2020]. Despite their complementary nature, selecting and interpreting the outputs of these different statistical tools remains challenging and requires a careful evaluation of their respective strengths and limitations when addressing ecological questions.

Predicting species interactions is a central goal in ecology [@valdovinos2019; @peralta2024], particularly in the context of climate change, ongoing species loss, and the need to sustain ecosystem services. However, although overall network structure can be partially reproduced [@eklof2013; @crea2016], accurately predicting pairwise interactions remains a major challenge [@valdovinos2019; @peralta2024]. Recent efforts to improve predictive performance have sought to address limitations of conventional analytical methods, such as non-independence of observations and non-linear effects of key variables, while incorporating the aforementioned ecological drivers [@bartomeus2016; @benadi2022]. In parallel, flexible machine learning methods have emerged as particularly promising for enhancing predictive accuracy [@strydom2021]. For example, @pichler2020 showed that machine learning models outperformed traditional regression approaches when predicting species interactions in plant-pollinator networks. Nevertheless, skepticism remains, and comparative approaches that evaluate predictive ability in different contexts are needed.

In this study, we aim to evaluate the relative importance of biotic and abiotic drivers of plant-pollinator interactions within seasonal dynamics using various analytical methods, and to assess the predictive power of species interactions under different analytical approaches and temporal scales (**Figure 1**). To achieve this, we collected data on plant-pollinator interactions, along with associated biotic (i.e., abundances, traits, and phenology) and abiotic variables (i.e., temperature), in three German botanical gardens. Our objective is to contribute to a mechanistic understanding of these drivers by using a semi-controlled environment with numerous co-flowering species that exhibit heterogeneous traits and diverse phylogenetic backgrounds. This setting allows us to test multiple potential combinations of pairwise interactions within the gardens and to explore in depth the role of traits and phylogenetic relatedness as surrogates for unmeasured traits.

\vspace{-2cm}


```{r, out.width="1\\linewidth", include=TRUE, fig.align="left", echo=FALSE}
knitr::include_graphics("../Data/figure1.pdf")
```

\vspace{-2.5cm}

**Figure 1**. Workflow of the study, showing how different ecological drivers (A) are tested across various temporal scales (B) using different analytical methods (C). Panel A illustrates the main ecological drivers used in this study to evaluate plant–pollinator interaction patterns: (1) pollinator and floral traits, (2) pollinator and floral abundances, and (3) pollinator and floral phenology. Panel B presents the three levels of temporal aggregation considered in this study: (1) individual sampling days, (2) early, mid, and late season, and (3) the full flowering season. Panel C depicts the three main analytical approaches implemented in this study to evaluate the effects of different ecological drivers across distinct levels of temporal aggregation: (1) matrix analyses, (2) generalized linear models (GLMs), and (3) machine learning


# Methods

## Study site and sampling method

Plant-pollinator interactions were monitored during the entire flowering season of 2023 in the botanical gardens of Halle, Jena, and Leipzig (Germany; **Figure S1**). These gardens belong to PhenObs initiative [@nordt2021], which consists of a network of botanical gardens across the Northern Hemisphere that records phenological observations on a large set of herbaceous species. The objective of PhenObs is to evaluate the impacts of climate change on species performance [e.g.\, @sporbert2022]. In this study, plant-pollinator interactions were recorded for a subset of species from the PhenObs project within these three gardens. Each garden was visited weekly, and plant-pollinator interactions were sampled during 15 mins on each focal plant.  On each sampling day, the different focal plants were sampled in a random order and only pollinator individuals that contacted the reproductive organs of the plant were recorded as interacting species. In addition, to create a comprehensive inventory of pollinators and their interactions within the gardens, 10 random sites were selected each sampling day, and plant-pollinator interactions were monitored for three minutes at each random site, resulting in a total of 30 minutes per sampling day and garden. Therefore, two sampling methodologies were implemented, one that recorded interactions on PhenObs plants, and another that consisted of random stops within the garden, with the specific goal of describing all potential pollinators present in the garden.

```{r ,echo=FALSE, message=FALSE, cache=FALSE, warning="FALSE"}

# Load libraries
library(dplyr)

# Load data
raw_data = readRDS("../Data/Working_files/interaction_data.rds")
phenobs_spp = readRDS("../Data/Working_files/phenobs_spp.rds")

# Number of phenobs species
phenobs_total = length(phenobs_spp)

# Prepare data
int_data = raw_data %>%
  filter(Plant_rank == "SPECIES") %>%
  select(Plant, 
         Plant_accepted_name, 
         Pollinator_accepted_name, 
         Botanical_garden, 
         Sampling,
         Plant_order,
         Plant_family,
         Plant_genus) %>%
  rename(Plants = Plant_accepted_name,
         Pollinators = Pollinator_accepted_name) %>%
  distinct()

# Check unique number of PhenObs species by garden
phenobs_by_garden = int_data %>% 
filter(Plant %in% phenobs_spp) %>% 
select(Botanical_garden, Plant) %>% 
group_by(Botanical_garden) %>% 
summarise(Phenobs_spp = n_distinct(Plant))

# PhenObs species by garden
phenobs_h = phenobs_by_garden$Phenobs_spp[1]
phenobs_j = phenobs_by_garden$Phenobs_spp[2]
phenobs_l = phenobs_by_garden$Phenobs_spp[3]

# Check unique number of Non-PhenObs species by garden
phenobs_by_garden1 = int_data %>% 
filter(Sampling == "Random_census") %>% 
select(Botanical_garden, Plant) %>% 
group_by(Botanical_garden) %>% 
summarise(Non_phenobs_spp = n_distinct(Plant))

# PhenObs species by garden
non_phenobs_h = phenobs_by_garden1$Non_phenobs_spp[1]
non_phenobs_j = phenobs_by_garden1$Non_phenobs_spp[2]
non_phenobs_l = phenobs_by_garden1$Non_phenobs_spp[3]

# Check total number of plant orders within PhenObs
phenobs_orders = int_data %>% 
  filter(Plant %in% phenobs_spp) %>% 
  distinct(Plant_order) %>% 
  nrow()

# Check total number of plant families within PhenObs
phenobs_families = int_data %>% 
  filter(Plant %in% phenobs_spp) %>% 
  distinct(Plant_family) %>% 
  nrow()

# Check total number of plant genera within PhenObs
phenobs_genera = int_data %>% 
  filter(Plant %in% phenobs_spp) %>% 
  distinct(Plant_genus) %>% 
  nrow()

# Calculate total time per garden for PhenObs plants
time_per_garden = raw_data %>% 
  filter(Sampling == "Focal") %>% 
  select(Botanical_garden, Date, Total_time_species, Plant) %>% 
  distinct() %>% 
  group_by(Botanical_garden) %>% 
  summarise(Total_time = sum(Total_time_species) / 60)

halle_hours = round(time_per_garden$Total_time[1], 1)
jena_hours = round(time_per_garden$Total_time[2], 1)
leipzig_hours = round(time_per_garden$Total_time[3],1)


# Calculate total time per garden for NON-PhenObs plants
time_per_garden1 = raw_data %>% 
  filter(Sampling == "Random_census") %>% 
  select(Botanical_garden, Date, Total_time_species, Plant) %>% 
  distinct() %>% 
  group_by(Botanical_garden) %>% 
  summarise(Total_time = sum(Total_time_species) / 60)

halle_hours_random = round(time_per_garden1$Total_time[1], 1)
jena_hours_random = round(time_per_garden1$Total_time[2], 1)
leipzig_hours_random = round(time_per_garden1$Total_time[3],1)


# Percentage of species with capitulum
capitulum_percent = int_data = raw_data %>%
  filter(Plant_rank == "SPECIES") %>%
  #filter(Sampling == "Focal") %>% 
  #filter(Plant %in% phenobs_spp) %>% 
  select(Plant, 
         Capitulum) %>% 
  distinct() %>% 
  group_by(Capitulum) %>% 
  summarise(n_plants = length(Plant) / nrow(.) * 100)

non_capitulum_percent = capitulum_percent %>% 
  filter(Capitulum == "No") %>% 
  pull(n_plants) %>% 
  round(1)

capitulum_percent = capitulum_percent %>% 
  filter(Capitulum == "Yes") %>% 
  pull(n_plants) %>% 
  round(1)


```

## Sampling effort and taxonomic coverage

This study monitored a total of `r phenobs_total` focal plant species that belonged to PhenObs initiative, which accounted for `r phenobs_genera` different genera, `r phenobs_families` families and `r phenobs_orders` different orders. In total, we sampled pollinators on `r phenobs_h` plants species in Halle (`r halle_hours` hours), `r phenobs_j` in Jena (`r jena_hours` hours) and `r phenobs_l` in Leipzig (`r leipzig_hours` hours), and only 12 species were shared across the three gardens. 

POLLINATORS



In addition, with the random stops across the garden we sampled a total of XXXX species that comprised XXXX genera, XXXXX families and XXXXX orders. This study accounted for a total of `r non_phenobs_h` plant species in Halle (`r halle_hours_random` hours), `r non_phenobs_j` in Jena (`r jena_hours_random` hours), and `r non_phenobs_l` in Leipzig (`r leipzig_hours_random` hours). Explain accumulation curves here!!!

FIGURE 2 can be S1 at some point



## Data description

To evaluate the different mechanisms that drive plant-pollinator interactions, detailed information on abundance, phenology and species traits for both plants and pollinators was recorded as follows: 

(i) Floral abundance was annotated for each focal plant species, and given that botanical gardens are a semi-control setting with relatively low abundance of pollinators, we were also able to annotate the different number of pollinator individuals on each sampled plant when documenting interactions. Despite not being fully independent of interactions, this approach offers a more realistic estimate of pollinator abundance than the standard method of estimating abundances from interactions, which can bias our understanding of the mechanisms driving network structure [@bluthgen2006; @vizentin2014]. Note that flowering species with tiny densely packed flowers were considered at inflorescence level (`r capitulum_percent`% of the total flowering species considered in this study). 


(ii) Given that plant phenology was already monitored on the PhenObs project, we have weekly phenological records with flowering percent for all focal plants. For 10 species with missing phenology data, we conducted generative additive models (GAM) with our records of flower number to build the phenology on the same sampling dates as the other species by using the *mgcv* package. Unlike plants, pollinator phenology relies on ‘stochastic’ interaction observations that may not represent their complete flying period. For this reason, we opted to complement our observational data from the Global Biodiversity Information Facility (GBIF). Germany is one of the countries with highest records of occurrence data [@rocha2021], and in particular, their cities (e.g., botanical gardens) contain most of those records [@sweet2022]. Thus, we opted to extract occurrence records around the three cities for the different pollinator species to complement the different pollinator phenologies. We first created a centroid within the 3 cities, and then, created a rectangular shape around it that expanded approximately 1000 kms in longitude and 500 kms in latitude (Figure S1). We used this shape because we expect environmental variables to be more consistent and phenology to be more similar across a longitudinal range than a latitudinal one. Thus, our download resulted in 2,933,772 occurrence records (https://doi.org/10.15468/dl.k5cnuf).

(iii) Plant reproductive traits were selected following the main traits driving the trait spectrum of flowering plants as shown in @lanuza2023 (i.e., flower number, plant height, style length, flower size, ovule number and selfing). In addition, we also included phenological traits (i.e., flower lifespan and flowering length), floral rewards (nectar volume and number of pollen grains per flower) and seed weight as a measurement of the reproductive investment (**Table S1a**). Pollinator traits included intertegular distance, body length and proboscis length (**Table S1b**). Both intertegular distance and body length were measured from images using ImageJ, and proboscis length was extrapolated for bee species by using the pollimetry package [@kendall2019]. Non-bee species were complemented with proboscis values from the literature and images when possible. 

## Plant and pollinator phylogeny


*Phylogeny... Plant phylogeny is easy, bee phylogeny too... 

*compute phenology matrices: presence absence and probability ones based by calculating the probability of observing both if they are independent events divided by total number of sampling days


CONCEPTUAL GRAPH. **Figure 1**

## Analysess

To unravel the role of time in our understanding of the cological drivers of plant-pollinator networks, this study analyses the relative importance of abundance, phenology and traits at the whole season level, by dividing in early-mid-and late season and by analysing at the sampling day level. 


# Results

PROVIDE: Percentage of forbidden links due to phenological mistmatches by garden.

# References

<div id="refs"></div>


# Supporting information

```{r, echo=FALSE, message=FALSE, warning=FALSE, out.width='60%', fig.align='center'}
library(gt)

library(gt)
library(webshot2)
library(knitr)

tab <- readRDS("../Data/Working_files/Trait_summary_table.rds")
tab <- tab |>
  tab_caption(
    md("**Table 1.** Summary of plant and pollinator traits.")
  )


fig_dir <- knitr::opts_chunk$get("fig.path")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

out <- file.path(fig_dir, "Trait_summary_table.png")
gtsave(tab, out, zoom = 13, vwidth = 3200)

stopifnot(file.exists(out))
knitr::include_graphics(out)
```

\newpage

Figure S1: MAP WITH POTENTIAL POINTS OF GBIF

\newpage

Figure S2: RAREFACTION CURVES

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="105%", fig.cap= ""}

# ======================================================
# Compute standardized accumulation curve
# ======================================================

# Load libraries
library(dplyr)
library(purrr)
library(ggplot2)
library(tibble)
library(minpack.lm)
library(patchwork)
library(ggnewscale)

# ======================================================
# Load data
raw_data <- readRDS("../Data/Working_files/interaction_data.rds")
colnames(raw_data)

# ======================================================
# 1) Prepare data to compute accumulation curves across species

# Get total sampling time per species and garden
sampling_times <- raw_data %>%
  filter(Plant_rank == "SPECIES") %>%
  select(Plant_accepted_name, Botanical_garden, Total_time_species, Date, Sampling) %>%
  distinct() %>%
  rename(Plants = Plant_accepted_name,
         Sampling_time = Total_time_species) %>% 
  group_by(Plants, Botanical_garden, Sampling) %>%
  summarise(Total_sampling_time = sum(Sampling_time), .groups = 'drop')

# Get unique interactions per garden
int_data <- raw_data %>%
  filter(Plant_rank == "SPECIES", 
         Pollinator_rank == "SPECIES") %>%
  select(Plant_accepted_name, Pollinator_accepted_name, Botanical_garden, Sampling) %>%
  rename(Plants = Plant_accepted_name,
         Pollinators = Pollinator_accepted_name) %>%
  distinct()

# Join data
data <- left_join(int_data, sampling_times) %>%
  arrange(Botanical_garden)

# ======================================================
# Accumulation curve function
curves <- function(data) {
  
  # Random order of all plants
  sampled_plant <- sample(unique(data$Plants), length(unique(data$Plants)))
  
  output <- data %>% 
    filter(Plants %in% sampled_plant) %>%
    arrange(match(Plants, sampled_plant)) %>%
    group_by(Pollinators) %>%
    mutate(distinct = row_number() == 1) %>%
    ungroup() %>%
    group_by(Plants) %>%
    mutate(Unique_spp = sum(distinct)) %>%
    ungroup() %>%
    select(Plants, Unique_spp, Botanical_garden, Sampling) %>%
    distinct() %>%
    mutate(Unique_spp_cumulative = cumsum(Unique_spp)) %>%
    mutate(Plant_sampled = row_number())
  
  output_time <- data %>%
    filter(Plants %in% sampled_plant) %>%
    arrange(match(Plants, sampled_plant)) %>%
    select(Plants, Total_sampling_time, Botanical_garden, Sampling) %>%
    distinct() %>%  # one row per plant
    mutate(
      Cumulative_time = cumsum(Total_sampling_time),
      Plant_sampled = row_number())
  
  output_curve <- left_join(output_time, output, by = c("Plants", "Plant_sampled",
                                                        "Botanical_garden","Sampling"))
  
  return(output_curve)
}

# ======================================================
# Run accumulation curves iteratively by garden and sampling type
results_by_garden_sampling <- data %>%
  group_split(Botanical_garden, Sampling) %>%
  map(~ map_dfr(1:100, function(i) curves(.x), .id = "iteration"))

iterations_combined <- bind_rows(results_by_garden_sampling)

# ======================================================
# Summarize mean curve per plant sampled, per garden and sampling type
summary_by_garden_sampling <- results_by_garden_sampling %>%
  map(~ .x %>%
        group_by(Plant_sampled, Botanical_garden, Sampling) %>%
        summarise(
          Mean_cumulative_spp = mean(Unique_spp_cumulative),
          Mean_cumulative_time = mean(Cumulative_time),
          .groups = "drop"))

# Combine summaries
summary_for_plot <- bind_rows(summary_by_garden_sampling)

# ======================================================
# Optional: Save data
# saveRDS(iterations_combined, "Data/Working_files/rarefied_curves.rds")
# saveRDS(summary_for_plot, "Data/Working_files/mean_rarefied_curve.rds")

# ======================================================
# Extrapolation to fixed max time
global_max_time <- 4000

fit_extrapolate_fixed_time <- function(df, max_time) {
  model <- nlsLM(Unique_spp_cumulative ~ a * (1 - exp(-(b * Cumulative_time)^c)),
                 data = df,
                 start = list(a = max(df$Unique_spp_cumulative), b = 0.001, c = 1))
  
  new_times <- data.frame(Cumulative_time = seq(0, max_time, length.out = 100))
  new_times$predicted <- predict(model, newdata = new_times)
  new_times$Botanical_garden <- unique(df$Botanical_garden)
  new_times$Sampling <- unique(df$Sampling)
  
  return(new_times)
}

# Apply extrapolation by group
extrapolated_data_fixed <- iterations_combined %>%
  group_by(Botanical_garden, Sampling) %>%
  group_split() %>%
  map_dfr(~ fit_extrapolate_fixed_time(.x, global_max_time))

# Max observed time per group
max_times_df <- iterations_combined %>%
  group_by(Botanical_garden, Sampling) %>%
  summarise(max_observed_time = max(Cumulative_time), .groups = 'drop')

# Add alpha group to extrapolated data
extrapolated_data_fixed <- extrapolated_data_fixed %>%
  left_join(max_times_df, by = c("Botanical_garden", "Sampling")) %>%
  mutate(alpha_group = ifelse(Cumulative_time <= max_observed_time, "observed", "extrapolated"))

# ======================================================

summary_for_plot_points = extrapolated_data_fixed %>% 
  filter(alpha_group == "observed") %>% 
  group_by(Botanical_garden, Sampling) %>% 
  summarise(Mean_cumulative_time = max(Cumulative_time),
            Mean_cumulative_spp = max(predicted))


# Final plot
ggplot() +
  # Iteration lines
  geom_path(data = iterations_combined, 
            aes(x = Cumulative_time, y = Unique_spp_cumulative, 
                group = interaction(iteration, Botanical_garden, Sampling),
                color = Plant_sampled),
            size = 0.3, alpha = 0.5) +
  scale_colour_gradient(low = "grey90", high = "grey10", guide = "colorbar") +
  labs(color = "Plants sampled") +
  new_scale_color() +
  
  # Mean extrapolated curves
  geom_line(data = extrapolated_data_fixed, 
            aes(x = Cumulative_time, y = predicted, 
                color = Botanical_garden, linetype = Sampling, alpha = alpha_group),
            size = 1.2) +
  scale_alpha_manual(values = c("observed" = 1, "extrapolated" = 0.5), guide = "none") +
  scale_colour_viridis_d() +
  scale_linetype_manual(
    values = c("Focal" = "solid", "Random_census" = 22),
    labels = c("Focal" = "Phenobs plants", "Random_census" = "Random plants")) +
    geom_point(data = summary_for_plot_points, 
             aes(x = Mean_cumulative_time, y = Mean_cumulative_spp, 
                 color = Botanical_garden, linetype = Sampling),
             size = 2.74) +
  scale_colour_viridis_d() +
  
  # Labels and theme
  xlab("Time (mins)") +
  ylab("Pollinators") +
  theme_minimal(base_size = 14) +
  labs(color = "Botanical Garden", linetype = "Sampling type", alpha = "Data Type")

#Read interaction daa
library(dplyr)
library(tidyr)
library(lubridate)
library(readr)
library(ggplot2)
library(viridis)
library(scales)
#Load data
raw_data = readRDS("../Data/Working_files/interaction_data.rds")


poll_species = raw_data %>% 
  filter(Pollinator_rank == "SPECIES") %>% 
  filter(Pollinator_order == "Hymenoptera" | Pollinator_order == "Diptera"| 
           Pollinator_order == "Coleoptera" |  Pollinator_order == "Lepidoptera") %>% 
  select(Pollinator_accepted_name, Pollinator_order) %>% 
  group_by(Pollinator_order) %>% 
  summarise(Species = n_distinct(Pollinator_accepted_name)/n_distinct(.),
            Interactions = length(Pollinator_accepted_name)/ nrow(.))

# Combine Interactions and Species into a single column
data_long = pivot_longer(poll_species, cols = c(Interactions, Species), names_to = "Variable", values_to = "Value")

#Set order of levels for plotting
data_long$Pollinator_order = factor(data_long$Pollinator_order, levels = c("Coleoptera", "Lepidoptera", "Diptera", "Hymenoptera"))

data_long$Variable = factor(data_long$Variable, levels = c("Species", "Interactions"))

p1 = ggplot(data_long, aes(x = Variable, y = Value, fill = Pollinator_order)) + 
  geom_bar(stat = "identity", width = 0.7, alpha=0.85) +
  theme_classic() +
  scale_fill_viridis(discrete=TRUE, direction = -1, name="Pollinator order") +
  xlab(NULL) +
  ylab(NULL) + 
  scale_y_continuous(breaks=c(0,0.5,1),labels = percent_format(), expand = c(0, 0)) +
  theme(axis.text.y = element_text(size=14), 
        axis.text.x = element_text(size=14), 
        axis.title.y = element_text(face = "bold", size = 18, vjust = -2),  
        axis.title.x = element_text(face="bold", size = 18),
        legend.position = "right",
        legend.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 14),
        legend.key.size = unit(5, "mm"),
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(-8,-5,-5,-5)) +
  guides(fill=guide_legend(nrow=2,byrow=TRUE))


p1


```
