---
title: "Relationships of water related leaf traits in the genus Anthurium"
format:
  pdf:
    include-in-header: 
      text: |
        \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
        \usepackage{caption}
        \captionsetup[figure]{labelformat=empty}
        \captionsetup[table]{labelformat=empty}
        \usepackage{float}
        \floatplacement{figure}{H}

latex-tinytex: false
link-citations: TRUE
linkcolor: RoyalBlue
urlcolor: RoyalBlue
---
```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,}
#Load libraries
#Load libraries
library(dplyr) #data organization
library(vegan) #nmds
library(tidyr) #pivot wider
library(ggplot2) #plotting
library(viridis) #plotting
library(patchwork) #plotting
library(tibble) #cols to rows

```

# Project: PollObs (no title for now)

*Provisional author position:*

Jose B. Lanuza^1^, Henriette F. Morgenroth^1^,  Raymond Umazekabiri^1^, Marc Hoffmann^1^, Panagiotis Theodorou^1,2^, Robert Paxton^1,2^, Isabell Hensen^1,2^,  Robert Rauschkolb^1,3^, Christine Römermann^1,3^, Oliver Schweiger ^1,5^, Martin Freiberg^1,4^, Alfonso Allen-Perkins^6^, Tiffany Knight^1^

*Affiliations*

1. German Centre for Integrative Biodiversity Research (iDiv) Halle-JenaLeipzig, Puschstr. 4, 04103 Leipzig, Germany 
2. Institute for Biology, Martin-Luther-UniversityHalle-Wittenberg, Hoher Weg 8, 06120 Halle (Saale), Germany 
3. Institute of Ecology and Evolution with Herbarium Haussknecht and Botanical Garden, Friedrich Schiller University Jena, Jena, Germany 
4. Systematic Botany and Functional Biodiversity/Botanical Garden, Institute of Biology, Leipzig University, Johannisallee 21-23, 04103, Leipzig, Germany 
5. UFZ-Helmholtz Centre for Environmental Research, Department of Community Ecology, 06120 Halle, Germany
6. Departamento de Ingeniería Eléctrica, Electrónica, Automática y Física Aplicada, ETSIDI, Universidad Politécnica de Madrid, 28040 Madrid, Spain.

# Abstract

Flowering plants exhibit an astonishing diversity of reproductive structures, which  shapes how they interact with their pollinators. However, to what extent reproductive traits or other drivers like phenology and species abundances shape plant-pollinator interactions is still an open question. To disentangle the relative effects of species phenology, abundances and traits in plant-pollinator interactions, we compared pollinator visitation rates, diet breath, specialization from plant species that vary in reproductive traits, phenology and abundances. To do this, we observed plant-pollinator interactions for a whole flowering season in three different botanical gardens located in an urban setting. We anticipate that the ecological role of the different plan t and pollinator species are largely driven by their phenological overlap and abundance, and only certain taxonomic groups are strongly influenced by trait matching processes. Our results highlight the relevance of considering multiple mechanisms to understand plant-pollinator interactions.


# Main objective

Explore how traits, phenology and abundances shape species interactions in three botanical gardens from Germany. 

Pending tasks:

- Clean trait and phenology data for plants

- Get Pollinator phenology and traits

- Easy things to check: alpha, beta and gamma diversity of pollinators across the three gardens.

# Methods

WRITE XXXXX

# The data has an important temporal component

In Figure 1 we can see in a color coded NMDS the strong temporal component of the composition across gardens and different taxonomic levels.

```{r ,echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, fig.height=8,fig.width=8, fig.cap = "\\textbf{Figure 1}. Non-metric multidimensional scaling (NMDS) of the pollinator community composition by sampling day in the three gardens at family, genus and species level."}

#Script to check how the pollinator community changes throughout the season
#Ordination analyses-->NMDS
#Load data
data = readRDS("../Data/Working_files/interaction_data.rds")
#Select columns of interest
d = data %>% 
select(Botanical_garden, Pollinator_rank, Pollinator_accepted_name,
       Pollinator_family, Pollinator_genus, Date_time) %>% 
mutate(Date = as.Date(Date_time)) %>% 
mutate(Date = format(as.POSIXct(Date, format="%Y%m%d"),format="%d%b%y")) %>% 
select(!Date_time) %>% 
rename(Family = Pollinator_family) %>% 
rename(Genus = Pollinator_genus) %>% 
rename(Species = Pollinator_accepted_name)

#Create function to run for each garden and taxonomic level
nmds_plot = function(data, garden, taxanomic_level){
#1)Pollinator family level
d1 = d %>% 
filter(Botanical_garden == garden) %>% 
select(!!sym(taxanomic_level), Date) %>% 
filter(!is.na(!!sym(taxanomic_level))) %>% 
group_by(Date, !!sym(taxanomic_level)) %>% 
count(name = "Number")
#Convert to wide and then to matrix
matrix = d1 %>% 
pivot_wider(names_from = !!sym(taxanomic_level), 
            values_from = Number, 
            values_fill = 0) %>%
remove_rownames %>% 
column_to_rownames(var="Date") %>%
as.matrix()
#Ordination analysis
nmds_output = metaMDS(matrix, 
                     k=2, trace=FALSE) # The number of reduced dimensions
#Visualize with ggplot
nmds_output_data = nmds_output$points %>%
as_tibble(rownames = "Date") %>%
mutate(Date = as.Date(format(as.POSIXct(Date, format="%d%b%y"), format="%Y-%m-%d"))) 
#Plot it
plot_nmds = ggplot(nmds_output_data, aes(MDS1,MDS2)) +
geom_point(stroke=1, aes(color=Date),size=2) + 
scale_colour_viridis(name = "Flowering",
                     labels=c("Start", "End"),
                    # labels=function(x)as.Date(x, origin="1970-01-01"),
                     breaks = c(min(nmds_output_data$Date), max(nmds_output_data$Date))) +
theme_bw() +
ggtitle(label = garden,
              subtitle = taxanomic_level)

plot_nmds
}

#Leipzig
l1 = nmds_plot(d, "Leipzig", "Family")
l2 = nmds_plot(d, "Leipzig", "Genus") + ggtitle(NULL)
l3 = nmds_plot(d, "Leipzig", "Species") + ggtitle(NULL)
#Halle
h1 = nmds_plot(d, "Halle", "Family")
h2 = nmds_plot(d, "Halle", "Genus") + ggtitle(NULL)
h3 = nmds_plot(d, "Halle", "Species") + ggtitle(NULL)
#Jena
j1 = nmds_plot(d, "Jena", "Family")
j2 = nmds_plot(d, "Jena", "Genus") + ggtitle(NULL)
j3 = nmds_plot(d, "Jena", "Species") + ggtitle(NULL)

#Family level
family_panel = j1 + h1 + l1   & theme(legend.position = "none", plot.title = element_text(size=16, face= "bold", hjust = 0.5))
#Genus level
genus_panel = j2 + h2 + l2   + plot_layout(guides = "collect") & theme(plot.title = element_text(size=16, face= "bold", hjust = 0.5))
#Species level
species_panel = j3 + h3 + l3   & theme(legend.position = "none", plot.title = element_text(size=16, face= "bold", hjust = 0.5))

family_panel /
genus_panel /
species_panel 


```

\newpage

# Plant phenologies

The phenology of plants and pollinators was monitored weekly on each of the three gardens. Unlike plants, pollinator phenology relies on 'stochastic' interaction obervations that may not represent their complete flying period. Thus, we built a comprehensive flying period of all pollinators by retrieving GBIF records for all observed floral visitors during the sampling year. In addition, this download was limited by area, where we considered a rectangular shape that extends more on the longitude range than the latitude one, as we expect more homogeneous environmental conditions in the former (**Figure X**). 



```{r ,echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, fig.height=10,fig.width=10, fig.cap = "\\textbf{Figure 2}."}


#Plot plant phenology together in this script

#Load libraries
library(dplyr)
library(ggplot2)

#Read botanical garden data
jena_phen = readRDS("../Data/Phenology_data/clean_plant_phenobs_jena.rds")
halle_phen = readRDS("../Data/Phenology_data/clean_plant_phenobs_halle.rds")
leipzig_phen = readRDS("../Data/Phenology_data/clean_plant_phenobs_leipzig.rds")

#Jena----
v = unique(jena_phen$Species)
d1 = jena_phen %>% 
filter(Species %in% v) %>% 
mutate(Flowering_intensity = if_else(Flowers_opening=="no", 0, Flowering_intensity)) 
#Find first numeric value of flowering intensity to order species
lev_species = d1 %>% 
group_by(Species) %>% 
filter(Flowers_opening == "y") %>% 
slice_min(Doy) %>% 
arrange(Doy) %>% 
pull(Species)
#This should be the order to be printed (from 1st flowering to last)
d1$Species = factor(d1$Species, levels = lev_species)
#Generate as many colors as species
colfunc = colorRampPalette(c("cyan4", "brown3"))
cols = colfunc(nlevels(d1$Species))


#Edit this
label_data = d1 %>% 
group_by(Species) %>% 
filter(Flowers_opening == "y") %>% 
slice_min(Doy) %>% 
arrange(Doy) %>% 
mutate(Flowering_intensity = 0) %>% 
select(Species, Doy, Flowering_intensity) %>% 
mutate(Doy = Doy-30)


#Plot
jena_plot = d1 %>% 
ggplot(aes(x = Doy, y = Flowering_intensity, fill = Species)) + 
geom_text(data = label_data, aes(label = Species), fontface="italic",
            vjust = -0.3, hjust = -0.1, size = 3, color = "black") +
stat_smooth(method = "gam",
method.args=list(family=poisson),
geom = "area",
alpha=0.75,
span = 0.1) +
theme_minimal()+
theme(legend.position = "none", 
      strip.background = element_blank(),
      strip.text.x = element_blank(),
      panel.spacing = unit(-1.5,'lines')) +
coord_cartesian(expand = FALSE) +
scale_y_continuous(expand = c(0,0), breaks = c(NULL), labels = c(NULL)) +
facet_wrap(~Species, ncol=1) +
ylab("Flowering intensity") +
xlab("Day of the year") +
scale_fill_manual(values = cols)  +
ggtitle("Jena")

#Halle----
v = unique(halle_phen$Species)
d1 = halle_phen %>% 
filter(Species %in% v) %>% 
mutate(Flowering_intensity = if_else(Flowers_opening=="no", 0, Flowering_intensity)) 
#Find first numeric value of flowering intensity to order species
lev_species = d1 %>% 
group_by(Species) %>% 
filter(Flowers_opening == "y") %>% 
slice_min(Doy) %>% 
arrange(Doy) %>% 
pull(Species)
#This should be the order to be printed (from 1st flowering to last)
d1$Species = factor(d1$Species, levels = lev_species)
#Generate as many colors as species
colfunc = colorRampPalette(c("cyan4", "brown3"))
cols = colfunc(nlevels(d1$Species))
#Edit this
label_data = d1 %>% 
group_by(Species) %>% 
filter(Flowers_opening == "y") %>% 
slice_min(Doy) %>% 
arrange(Doy) %>% 
mutate(Flowering_intensity = 0) %>% 
select(Species, Doy, Flowering_intensity) %>% 
mutate(Doy = Doy-30)

#Plot
halle_plot = d1 %>% 
ggplot(aes(x = Doy, y = Flowering_intensity, fill = Species)) + 
geom_text(data = label_data, aes(label = Species), fontface="italic",
            vjust = -0.3, hjust = -0.1, size = 3, color = "black") +
stat_smooth(method = "gam",
method.args=list(family=poisson),
geom = "area",
alpha=0.75,
span = 0.1) +
theme_minimal()+
theme(legend.position = "none", 
      strip.background = element_blank(),
      strip.text.x = element_blank(),
      panel.spacing = unit(-1.5,'lines')) +
coord_cartesian(expand = FALSE) +
scale_y_continuous(expand = c(0,0), breaks = c(NULL), labels = c(NULL)) +
facet_wrap(~Species, ncol=1) +
ylab(NULL) +
xlab("Day of the year") +
scale_fill_manual(values = cols)  +
ggtitle("Halle")

##Leipzig----
v = unique(leipzig_phen$Species)
d1 = leipzig_phen %>% 
filter(Species %in% v) %>% 
mutate(Flowering_intensity = if_else(Flowers_opening=="no", 0, Flowering_intensity)) 
#Find first numeric value of flowering intensity to order species
lev_species = d1 %>% 
group_by(Species) %>% 
filter(Flowers_opening == "y") %>% 
slice_min(Doy) %>% 
arrange(Doy) %>% 
pull(Species)
#This should be the order to be printed (from 1st flowering to last)
d1$Species = factor(d1$Species, levels = lev_species)
#Generate as many colors as species
colfunc = colorRampPalette(c("cyan4", "brown3"))
cols = colfunc(nlevels(d1$Species))
#Edit this
label_data = d1 %>% 
group_by(Species) %>% 
filter(Flowers_opening == "y") %>% 
slice_min(Doy) %>% 
arrange(Doy) %>% 
mutate(Flowering_intensity = 0) %>% 
select(Species, Doy, Flowering_intensity) %>% 
mutate(Doy = Doy-30)

#Plot
leipzig_plot = d1 %>% 
ggplot(aes(x = Doy, y = Flowering_intensity, fill = Species)) + 
geom_text(data = label_data, aes(label = Species), fontface="italic",
            vjust = -0.3, hjust = -0.1, size = 3, color = "black") +
stat_smooth(method = "gam",
method.args=list(family=poisson),
geom = "area",
alpha=0.75,
span = 0.1) +
theme_minimal()+
theme(legend.position = "none", 
      strip.background = element_blank(),
      strip.text.x = element_blank(),
      panel.spacing = unit(-1.5,'lines')) +
coord_cartesian(expand = FALSE) +
scale_y_continuous(expand = c(0,0), breaks = c(NULL), labels = c(NULL)) +
facet_wrap(~Species, ncol=1) +
ylab(NULL) +
xlab("Day of the year") +
scale_fill_manual(values = cols)  +
ggtitle("Leipzig")

#Order Jena-Halle-Leipzig
library(patchwork)
jena_plot + halle_plot + leipzig_plot & theme(axis.title = element_text(size=16, face= "bold"), plot.title = element_text(size=20, face= "bold", hjust = 0.5))



```
